# BMEG 310 Final Project 
Group 31: Monesha Basker, Gaurab Acharya, Ellie Smith

## Part 0 Loading Libraries
```{r}
library(ggplot2)
library(dplyr)
library(stringr)
```

## Part 0.1 Loading data 
```{r}
clinical_data <- read.table("Data/data_clinical_patient.txt", 
                            header = TRUE, sep = "\t", stringsAsFactors = FALSE)

mutation_data <- read.table("Data/data_mutations.txt", 
                            header = TRUE, sep = "\t", stringsAsFactors = FALSE)

expression_data <- read.csv("Data/RNAseq_KIRC.csv", 
                            header = TRUE, stringsAsFactors = FALSE)
```

## 0.2 Getting Patients that are in all data
```{r}
# Unique clinical patient ids
clinical_patient_ids <- unique(clinical_data$PATIENT_ID)

# Getting unique patient ids in mutation data
patients_in_mutation <- mutation_data$Tumor_Sample_Barcode
unique_patients_in_mutation <- unique(patients_in_mutation)
for (i in 1:length(unique_patients_in_mutation)) {
  string <- unique_patients_in_mutation[i]
  split_string <- unlist(strsplit(string, "-"))
  result <- paste(split_string[1:3], collapse = "-")
  unique_patients_in_mutation[i] <- result
}

# Getting unique patient ids in rnaseq data
patients_in_rnaseq <- colnames(expression_data)
patients_in_rnaseq <- patients_in_rnaseq[-1]
unique_rnaseq_patients <- list()
for (string in patients_in_rnaseq) {
  split_string <- unlist(strsplit(string, "\\."))
  result <- paste(split_string[1:3], collapse = ".")
  result <- gsub("\\.", "-", result)
  if (!(result %in% unique_rnaseq_patients)) {
    unique_rnaseq_patients <- append(unique_rnaseq_patients, result)
  }
}

common_patient_ids <- list()

for(patient in clinical_patient_ids) {
  if ((patient %in% unique_patients_in_mutation) &
      (patient %in% unique_rnaseq_patients)) {
    common_patient_ids <- append(common_patient_ids, patient)
  }
}

```

## 0.3 Changing Dataset to include patient id's 
```{r}
# Mutation data with Patient ID's
mutation_data <- mutation_data %>% 
  mutate(PATIENT_ID = str_extract(Tumor_Sample_Barcode, "^[^-]+-[^-]+-[^-]+"))

# Creating New rna_seq data 
patients_in_rnaseq <- colnames(expression_data)
patients_in_rnaseq <- patients_in_rnaseq[-1]
  
processed_colnames <- patients_in_rnaseq %>%
  sapply(function(string) {
    split_string <- unlist(strsplit(string, "\\."))
    result <- paste(split_string[1:3], collapse = ".")
    gsub("\\.", "-", result)  # Replace "." with "-"
  })

# New Expression / rna_seq aata with columns as Patient Id's
rna_seq_data <- expression_data
colnames(rna_seq_data)[-1] <- processed_colnames
```

## 0.4 Explore Clinical Data for Survival Analysis 
```{r}
table(clinical_data$OS_STATUS)
table(clinical_data$AJCC_PATHOLOGIC_TUMOR_STAGE)
table(clinical_data$PERSON_NEOPLASM_CANCER_STATUS)

filtered_data <- clinical_data %>%
  filter(clinical_data$PATIENT_ID %in% common_patient_ids)


ggplot(filtered_data, aes(x = "Cases", y = AGE)) +
  geom_boxplot(fill = "salmon") +
  labs(title = "Age Distribution of Cases with Complete Data", y = "Age") +
  theme_minimal()

non_synonymous_data <- mutation_data %>%
  filter(Variant_Classification %in% c("Missense_Mutation", "Nonsense_Mutation",
                                       "Frame_Shift_Del", "Frame_Shift_Ins"))
patients_in_mutation <- non_synonymous_data$Tumor_Sample_Barcode

for (i in 1:length(patients_in_mutation)) {
  string <- patients_in_mutation[i]
  split_string <- unlist(strsplit(string, "-"))
  result <- paste(split_string[1:3], collapse = "-")
  patients_in_mutation[i] <- result
}

non_synonymous_data$Patient_ID <- patients_in_mutation

mutation_counts <- non_synonymous_data %>%
  group_by(Patient_ID) %>%
  summarise(non_syn_mutations = n()) %>%
  filter(Patient_ID %in% common_patient_ids)

ggplot(mutation_counts, aes(x = Patient_ID, y = non_syn_mutations)) +
  geom_bar(stat = "identity", fill = "salmon") +
  labs(title = "Non-synonymous Mutations per Patient", 
       x = "Patient ID", y = "Number of Non-synonymous Mutations") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

# 1.0 Clinical Data Analysis 

## 1.1 Mutate Clinical Data by changing char columns to int boolean expressions
```{r}
# DSS -> Dead or alive without tumor = 0, Deceased with Tumor = 1
# OS -> Alive = 0, Deceased = 1 
#DFS -> DeceaseFree = 0, Recurred/Progressed = 1
#PFS -> Censored = 0, Progression = 1 
clinical_data <- clinical_data %>% 
  mutate(DSS_STATUS_BOOL = as.numeric(str_split(DSS_STATUS, ":", simplify = TRUE)[, 1]),
         OS_STATUS_BOOL = as.numeric(str_split(OS_STATUS, ":", simplify = TRUE)[, 1]),
         DFS_STATUS_BOOL = as.numeric(str_split(DFS_STATUS, ":", simplify = TRUE)[, 1]),
         PFS_STATUS_BOOL = as.numeric(str_split(PFS_STATUS, ":", simplify = TRUE)[, 1])
         )
```

## 1.2 Get Required Data 
```{r}
clin_data <- clinical_data[clinical_data$PERSON_NEOPLASM_CANCER_STATUS 
                           == "With Tumor",
                           c("PATIENT_ID", 
                             "OS_STATUS",
                             "OS_MONTHS",
                             "DAYS_LAST_FOLLOWUP",
                             "AJCC_PATHOLOGIC_TUMOR_STAGE",
                             "SEX",
                             "DSS_STATUS",
                             "DSS_MONTHS",
                             "PFS_STATUS",
                             "PFS_MONTHS",
                             "GENETIC_ANCESTRY_LABEL",
                             "RACE"
                             )]

```

Survival Analysis Libraries
```{r}
library("survival")
library("survminer")
library("SummarizedExperiment")
```
Survival Analysis with OS_STATUS 
```{r}
clin_data$deceased = clin_data$OS_STATUS == "1:DECEASED"

clin_data$overall_survival = ifelse(clin_data$deceased,
                                   clin_data$OS_MONTHS,
                                   clin_data$DAYS_LAST_FOLLOWUP)

columns_to_analyze <- list(clin_data$AJCC_PATHOLOGIC_TUMOR_STAGE,
                        clin_data$SEX,
                        clin_data$GENETIC_ANCESTRY_LABEL,
                        clin_data$RACE)
# Tumor Stage
fit <- survfit(Surv(overall_survival, deceased) ~ AJCC_PATHOLOGIC_TUMOR_STAGE,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Tumor Stage")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "OS-STATUS_","Tumor-Stage", ".png"),
       plot = plot$plot, bg = "white")
# Sex
fit <- survfit(Surv(overall_survival, deceased) ~ SEX,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Sex")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "OS-STATUS_","Sex", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

# Genetic Ancestry 
fit <- survfit(Surv(overall_survival, deceased) ~ GENETIC_ANCESTRY_LABEL,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Genetic Ancestry")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "OS-STATUS_","Genetic-Ancestry", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

# Race 
fit <- survfit(Surv(overall_survival, deceased) ~ RACE,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Race")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "OS-STATUS_","Race", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

```
Survival Analysis with DSS_STATUS 
```{r}
clin_data$deceased = clin_data$DSS_STATUS == "1:DEAD WITH TUMOR"

clin_data$overall_survival = ifelse(clin_data$deceased,
                                   clin_data$DSS_MONTHS,
                                   clin_data$DAYS_LAST_FOLLOWUP)

# Tumor Stage
fit <- survfit(Surv(overall_survival, deceased) ~ AJCC_PATHOLOGIC_TUMOR_STAGE,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Tumor Stage")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "DSS-STATUS_","Tumor-Stage", ".png"),
       plot = plot$plot, bg = "white")
# Sex
fit <- survfit(Surv(overall_survival, deceased) ~ SEX,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Sex")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "DSS-STATUS_","Sex", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

# Genetic Ancestry 
fit <- survfit(Surv(overall_survival, deceased) ~ GENETIC_ANCESTRY_LABEL,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Genetic Ancestry")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "DSS-STATUS_","Genetic-Ancestry", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

# Race 
fit <- survfit(Surv(overall_survival, deceased) ~ RACE,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Race")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "DSS-STATUS_","Race", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

```

Survival Analysis with PFS_STATUS 
```{r}
clin_data$deceased = clin_data$PFS_STATUS == "1:PROGRESSION"

clin_data$overall_survival = ifelse(clin_data$deceased,
                                   clin_data$PFS_MONTHS,
                                   clin_data$DAYS_LAST_FOLLOWUP)

# Tumor Stage
fit <- survfit(Surv(overall_survival, deceased) ~ AJCC_PATHOLOGIC_TUMOR_STAGE,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Tumor Stage")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "PFS-STATUS_","Tumor-Stage", ".png"),
       plot = plot$plot, bg = "white")
# Sex
fit <- survfit(Surv(overall_survival, deceased) ~ SEX,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Sex")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "PFS-STATUS_","Sex", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

# Genetic Ancestry 
fit <- survfit(Surv(overall_survival, deceased) ~ GENETIC_ANCESTRY_LABEL,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Genetic Ancestry")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "PFS-STATUS_","Genetic-Ancestry", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")

# Race 
fit <- survfit(Surv(overall_survival, deceased) ~ RACE,
               data=clin_data)
plot <- ggsurvplot(
  fit,
  data = clin_data,
  pval = TRUE,
  risk.table = TRUE,         
  risk.table.col = "strata",
  risk.table.height = 0.4,  
  ggtheme = theme_minimal(), 
  title = paste("Survival Analysis by Race")
)
print(plot)
ggsave(filename = paste0("survivalplots/", "PFS-STATUS_","Race", ".png"),
       plot = plot$plot, width = 8, height = 6, bg = "white")


```
